Abstractions for Multi-Sorted Substitutions

### 2023-09-16 paper/paper.tex
Update.

\begin{document}
  \begin{abstract}
    Formalizing a typed programming language in a proof assistant
    requires to choose representations for variables and typing.
    Variables are often represented as de Bruijn indices, where
    substitution is usually defined in terms of renamings to allow for
    proofs by structural induction.
    Typing can be represented extrinsically by defining untyped terms and a typing
    relation, or intrinsically by combining syntax and typing into a single
    definition of well-typed terms.
    For extrinsic typing, there is again a choice between extrinsic
    scoping, where terms and the notion of free variables are defined
    separately, and intrinsic scoping, where terms are indexed by their
    free variables.

@@

  % \begin{abstract}
  %   Formalizing a typed programming language in a proof assistant
  %   requires to choose between an extrinsic and an intrinsic
  %   representation of typing.
  %   \emph{Extrinsic} typing means to separately define a set of
  %   untyped terms and a typing relation.
  %   \emph{Intrinsic} typing means to directly define a set of well-typed terms,
  %   incorporating the typing relation into the syntax definition.

  %   One consequence of intrinsic typing is that operations on terms
  %   are also proofs that the operations preserve typing.
  %   This often leads to less code, as shared structure between
  %   definitions and preservation lemmas does not need to be repeated,
  %   but has the drawback that specification and proofs are
  %   intertwined, making it more difficult to check if a specification is
  %   correct.
  % \end{abstract}

  % \begin{abstract}
  %   Formalizing a typed programming language in a proof assistant
  %   requires to choose between an extrinsic and intrinsic representation of typing.
  %   For extrinsic typing, 
  %   There are two approaches to formalizing the syntax of typed object languages
  %   in proof assistants. The extrinsic approach defines
  %   object terms and typing relation separately, whereas the
  %   instrinsic approach only defines well-typed terms.

  %   This paper describes an Agda framework for deriving substitution
  %   operations and lemmas for syntaxes with extrinsic typing,
  %   intrinsic scoping, and DeBruijn Indices as variable representation.

  %   This paper describes an Agda framework for formalizing programming
  %   languages with extrinsic typing, intrinsic scoping, and DeBruijn
  %   Indices as variable representation. The key feature of the
  %   framework 
  %   all sorts of object variables.
  %   Agda's reflection mechanism is used to derive 
  %   substitution properties for a
  %   large class of syntaxes and only requires user code for to prove that the

  %   framework is to deal with multiple sorts of variables in a uniform way
  % \end{abstract}

  % \begin{abstract}
  %   Variable binding and substitution are central concepts in the formalization
  %   of programming languages in proof assistants. 
  %   % Many different representations of variables have been developed with different
  %   % trade-offs on proof simplicity, clarity, and reduction of boilerplate.
  %   A common approach is to represent variables as DeBruijn Indices
  %   and to define substitution in terms of variable renamings for the sake
  %   of structural induction.
  % \end{abstract}

  % \begin{abstract}
  %   Variable binding and substitution are a long-standing nuisance for mechanized
  %   metatheory of programming languages. A common approach is to represent
  %   variables as DeBruijn Indices, where substitution is defined in terms of
  %   variable renaming. If a language has $n$ different kinds of variables, and
  %   $m$ different kinds of terms, we potentially need to define both
  %   renaming and substitution for all $m\cdot n$ combinations and prove the
  %   theory of those operations both in isolation and with respect to each other.

  %   % McBride introduced kits to define both substitution and renaming by a
  %   % single term traversal for intrinsically typed terms.\cite{mcbride2005kits}

  %   We propose a technique to derive the implementation and metatheory of
  %   renaming and substitution for all $m\cdot n$ combinations of variables and terms
  %   from a single term traversal and three simple lemmas for extrinsically typed
  %   but intrinsically scoped terms.
  %   We demonstrate this technique by formalizing type soundness of System F with
  %   very parallel substitutions, which replace both type and term variables at
  %   once, enabling a single substitution-preserves-typing lemma.
  %   We provide a well-documented implementation of the technique as an Agda
  %   library.

  %   As the technique is based on extrinsic typing, the metatheory of renaming
  %   and substitution is decoupled from type preservation.
  %   This has the downside, that type preservation of renaming and substitution
  %   are not for free, but also the upside, that the library can be directly
  %   instantiated for object languages with non-standard typing relations like
  %   dependent types, type state, or resource environments.
  % \end{abstract}

@@

  \section{Introduction}
  \label{sec:introduction}
  In recent years intrinsic typing\cite{TODO} has become rather popular.
  A consequence of intrinsic typing is that operations on terms
  are also proofs that the operations preserve typing.
  This often has the benefit of more concise code, as shared structure
  between definitions and preservation lemmas does not need to be
  repeated.
  However, intrinsic typing also has three major drawbacks:
  \begin{enumerate}
  \item
    It is more difficult to judge whether a specification is correct,
    because there is no clear separation between specification and proofs.
    To understand the specification one has to additionally
    understand the proofs, which can introduce significant cognitive overhead.
    % For example specifying a small step semantics is directly intertwined
    % with the proof of subject reduction.
  \item
    It is less uniform for object languages with multiple variable
    sorts, e.g. type- and expression-variables. In such a scenario,
    types are usually indexed by kinds and expressions by kinds and types.
    This requires not only a type-in-type and expression-in-expression
    substitution, but also a type-in-expression substitution, which causes
    interactions between different types of substitutions introducing
    additional complexity.
    The duplication increases even more if one additionally introduces
    kind-variables. Then a total of 6 different substitutions and
    their interactions is required.
    Unifying these substitutions to reduce repetition is difficult
    with intrinsic typing due to the interdependence of definitions.
  \item
    It is less suitable for automation, because type preservation is baked
    into the syntax definition and type preservation can become arbitrary complex
    depending on the typing relation. This can prevent substitition
    definition and lemmas from being derived, where otherwise it would be possible.
  \end{enumerate}
  With extrinsic typing the major drawback is the duplication between
  definitions and type preservation lemmas, which we found more
  suitable to be eliminated by automation.

@@

  \subsection{Structure}
  \label{sec:introduction:structure}
  The rest of this paper will introduce our framework using System F
  as a running example for a type soundness proof.

>>>>

\begin{document}

  \begin{abstract}
    Formalizing a typed programming language in a proof assistant
    requires to choose representations for variables and typing.
    Variables are often represented as de Bruijn indices, where
    substitution is usually defined in terms of renamings to allow for
    proofs by structural induction.
    Typing can be represented extrinsically by defining untyped terms and a typing
    relation, or intrinsically by combining syntax and typing into a single
    definition of well-typed terms.
    For extrinsic typing, there is again a choice between extrinsic
    scoping, where terms and the notion of free variables are defined
    separately, and intrinsic scoping, where terms are indexed by their
    free variables.

@@

@@

  \section{Introduction}
  \label{sec:introduction}

  Formalizing a typed programming language in a proof assistant
  requires to choose between an extrinsic and intrinsic representation
  of typing.

  With \emph{intrinsic typing}, the definitions of syntax and typing
  are merged into a single definition of well-typed terms.
  Consequently, operations on terms have to map
  well-typed terms to well-typed terms proving that they preserve typing.
  For example, in an intrinsic setting, the definition of substitution
  also acts as a proof that substitution preserves typing, and the
  definition of a small-step semantics also acts as a proof for subject
  reduction.
  When working without library support, this often has the benefit of
  more concise code, as shared structure between definitions and
  preservation lemmas does not need to be repeated.
  However, intrinsic typing also has three major drawbacks:
  \begin{enumerate}
  \item
    It is more difficult to judge whether a specification is correct,
    because there is no clear separation between specification and proofs.
    To understand the specification one has to additionally
    understand the proofs, which can introduce significant cognitive overhead.
    % For example specifying a small step semantics is directly intertwined
    % with the proof of subject reduction.
  \item
    It is less uniform for object languages with multiple variable
    sorts, e.g. type- and expression-variables. In such a scenario,
    types are usually indexed by kinds and expressions by kinds and types.
    This requires not only a type-in-type and expression-in-expression
    substitution, but also a type-in-expression substitution, which causes
    interactions between different types of substitutions introducing
    additional complexity.
    The duplication increases even more if one additionally introduces
    kind-variables. Then a total of 6 different substitutions and
    their interactions is required.
    Unifying these substitutions to reduce repetition is difficult
    with intrinsic typing due to the interdependence of definitions.
  \item
    It is less suitable for automation, because type preservation is baked
    into the syntax definition and type preservation can become arbitrary complex
    depending on the typing relation. This can prevent substitition
    definition and lemmas from being derived, where otherwise it would be possible.
  \end{enumerate}
  With extrinsic typing the major drawback is the duplication between
  definitions and type preservation lemmas, which we found more
  suitable to be eliminated by automation.

@@

  \subsection{Structure}
  \label{sec:introduction:structure}

  The rest of this paper will introduce our framework using System F
  as a running example for a type soundness proof.
